Illustrating Model Properties

### 2024-06-07 manuscript.qmd
split ipsi scenarios by midext
risk with fancy legend

![Comparison of predicted (histograms) vs observed (beta posteriors) prevalences. Shown are four scenarios, all including contralateral LNL II involvement: Early T-category and an ipsilateral N0 neck (green), early T-category and ipsilateral LNL II involvement (blue), as well as the same two scenarios but for advanced T-category (orange and red). The figure shows that the model is capable of describing the correlation between ipsi- and contralateral involvement. Although for the scenario of LNL II involvement in both sides, the prediction's split between early and advanced T-category is not large enough.](figures/fig-model-prevalences-with-ipsi.png){#fig-model-prevalences-with-ipsi width="50%"}

@@

- compute risks for occult disease, given diagnosis

>>>>

![Comparison of predicted (histograms) vs observed (beta posteriors) prevalences. Shown are four scenarios, all including contralateral LNL II involvement: Early T-category and an ipsilateral N0 neck (green), early T-category and ipsilateral LNL II involvement (blue), as well as the same two scenarios but for advanced T-category (orange and red). The figure shows that the model is capable of describing the correlation between ipsi- and contralateral involvement. Although for the scenario of LNL II involvement in both sides, the prediction's split between early and advanced T-category is not large enough.](figures/fig-model-prevalences-with-ipsi.png){#fig-model-prevalences-with-ipsi}

@@

In this section, we briefly investigate how the model may be applied clinically: We want to estimate the risk for occult disease in some or all LNLs, given the patient's individual diagnosis. In terms of our model, this diagnosis consists of the T-category, the lateralization of the tumor (does it extend over the mid-sagittal line?), and which LNLs are clinically involved, e.g. because some lymph nodes appear enlarged in an MRI scan or show increased glycolytic activity on an FDG PET/CT scan.

![Histograms over the predicted risk of contralateral LNL II invovlement, shown for all eight possible combinations of the risk factors T-category (early: top row, advanced: bottom row), midline extension (absent: left column, present: right column), and ipsilateral diagnosis (ipsilateral N0: blue, ipsilateral at least LNL III diagnosed: orange). The provided diagnosis always included a clinically node negative contralateral neck. Each of the three risk factors significantly increase the predicted risk for occult disease](figures/fig-model-risks.png){#fig-model-risks}

### 2024-06-07 model.qmd
risk with fancy legend

>>>>

# Risk Predictions

Now, we want to make actual use of the model and predict the risk for occult diease given some diagnoses.

```{python}
import numpy as np
import matplotlib.pyplot as plt
from matplotlib.legend_handler import HandlerTuple
from lyscripts.scenario import Scenario
from lyscripts.plot.utils import Histogram, draw, COLORS, split_legends
from scripts import paths, shared

all_false = {lnl: False for lnl in ["I", "II", "III", "IV", "V"]}

plt.rcParams.update(shared.get_fontsizes())
plt.rcParams.update(shared.get_figsizes(width=17./2., aspect_ratio=2.))

fig, ax = plt.subplots()
ax.set_xlim(0., 15.)

contents = []
defaults = {
  "density": True,
  "histtype": "stepfilled",
  "alpha": 0.7,
  "bins": 50,
}
patches = {
  "early T-cat.": [],
  "advanced": [],
  "lateralized": [],
  "midline ext.": [],
  "ipsi N0": [],
  "≥ ipsi LNL III": [],
}

scenario = Scenario(
  t_stages=["early"],
  midext=False,
  diagnosis={
    "ipsi": {"CT": all_false},
    "contra": {"CT": all_false},
  },
  involvement={"contra": {"II": True}},
)
_v, bins, patch = Histogram.from_hdf5(
  filename=paths.model_dir / "full" / "risks.hdf5",
  dataname=scenario.md5_hash("risks"),
).draw(ax, hist=defaults)

patches["early T-cat."].append(*patch)
patches["lateralized"].append(*patch)
patches["ipsi N0"].append(*patch)

scenario = Scenario(
  t_stages=["late"],
  midext=False,
  diagnosis={
    "ipsi": {"CT": {"III": True}},
    "contra": {"CT": all_false},
  },
  involvement={"contra": {"II": True}},
)
_v, bins, patch = Histogram.from_hdf5(
  filename=paths.model_dir / "full" / "risks.hdf5",
  dataname=scenario.md5_hash("risks"),
).draw(ax, hist=defaults)

patches["advanced"].append(*patch)
patches["lateralized"].append(*patch)
patches["≥ ipsi LNL III"].append(*patch)

scenario = Scenario(
  t_stages=["early"],
  midext=True,
  diagnosis={
    "ipsi": {"CT": all_false},
    "contra": {"CT": all_false},
  },
  involvement={"contra": {"II": True}},
)
_v, bins, patch = Histogram.from_hdf5(
  filename=paths.model_dir / "full" / "risks.hdf5",
  dataname=scenario.md5_hash("risks"),
).draw(ax, hist=defaults)

patches["early T-cat."].append(*patch)
patches["midline ext."].append(*patch)
patches["ipsi N0"].append(*patch)

scenario = Scenario(
  t_stages=["late"],
  midext=True,
  diagnosis={
    "ipsi": {"CT": {"III": True}},
    "contra": {"CT": all_false},
  },
  involvement={"contra": {"II": True}},
)
_v, bins, patch = Histogram.from_hdf5(
  filename=paths.model_dir / "full" / "risks.hdf5",
  dataname=scenario.md5_hash("risks"),
).draw(ax, hist=defaults)

patches["advanced"].append(*patch)
patches["midline ext."].append(*patch)
patches["≥ ipsi LNL III"].append(*patch)

for label in patches:
  handle = tuple(patches[label])
  patches[label] = handle

ax.legend(
  handles=patches.values(),
  labels=patches.keys(),
  handler_map={tuple: HandlerTuple(ndivide=None, pad=0.2)},
  handlelength=2.2,
  handletextpad=0.4,
  ncols=3,
  labelspacing=0.3,
  columnspacing=1.,
)

plt.savefig(paths.figure_dir / "fig-model-risks.png", dpi=400)
plt.show()
```
